Cellular Automata and Computational Universality

### 2018-09-13 paper/paper.tex
Additions to paper

\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{subfig}
\usepackage{amsmath}
\usepackage{url}
\usepackage{diagbox}

@@

    \subsection{Self-organisation}
    Second of the characteristics exhibited by this cellular automaton is
    ``self-organisation'', whereby from an initial disorderly or chaotic state
    there appear ordered structures seemingly spontaneously. This is illustrated
    in Fig. \ref{fig:self-organisation} below: one can clearly see the emergence
    of similar triangular structures in the cone snail's shell, appearing from a
    chaotic background. This starting chaos is modelled in the cellular
    automaton by randomly assigning each cell a value of 0 or 1.

@@

    Reading the bottom row of this transition table, we get the binary number
    $01011010_{\text{bin}} = 90_{\text{dec}}$, allowing us to name this rule set
    Rule 90. Using this convention, we may now construct our own automata based
    solely on the knowledge of its name (at least for elementary cellular
    automata). For example, take Rule 110, introduced by Wolfram 1983
    \cite{WolframRule110}. We first convert $110_{\text{dec}}$ to binary, and
    then use this to fill in the table as above:

@@

    Two technically different automata may give rise to similar patterns, and
    indeed many automata are equivalent to each other thanks to simple
    transformations of theirunerlying geometry. The first of these
    transformations is reflection in the vertical axis; the result of applying
    this is known as the mirror rule \cite{elementary}. For example, take Rule
    30. Each time some neighbourhood produces a live cell in Rule 30, let's take
    this neighbourhood and reflect it in the vertical plane, giving the table
    below:

@@

    The first line of the table shows the ruleset for Rule 30, while the line
    below gives that of our new ruleset, made by applying the mirror rule to
    Rule 30. Thus we get Rule 30's mirror rule Rule $01010110_{\text{bin}}$ =
    Rule $86_{\text{dec}}$. Rules which generate the same ruleset under
    mirroring are called amphichiral, and of the 256 elementary cellular
    automata, 64 are so.

@@

    \subsection{Two dimensional cellular automata}
    Two dimensional cellular automata become a bit more interesting than their
    dimensionally-deficient brethren. Rather than having a single row of cells
    to consider at each generation, now a two-dimensional plane of cells is used
    to compute the evolution of the automaton. Whereas elementary automata
    looked only at the neighbourhood of $a_{i-1}, a_i,$ and $a_{i+1}$ when
    applying the set of rules, two-dimensional automata typically have a larger
    neighbourhood to consider. Two of the most common types of neighbourhood are
    the Moore neighbourhood and the von Neumann neighbourhood
    \cite{neighbourhoods}. The former includes the 8 cells directly surrounding
    the current cell being considered, while the latter only focuses on the four
    orthogonally ajacent cells. Ocassionally, one may use the extended von
    Neumann neighbourhood, for which the next four closest orthogonally adjacent
    cells are also included.

@@

    \subsection{Game of Life and Life-like automata}
    At this point it makes sense to talk a bit more about Conway's Game of Life.
    It was invented in 1970 by the British mathematician John Conway, after
    developing an interest in them after their introduction by John von Neumann
    in the late 1940s, who aimed to discover a hypothetical machine that was
    capable of recreating itself \cite{GameOfLife}. Von Neumann was able to
    achieve this goal using a complex automaton, with 29 states and complicated
    rules to do so \cite{VonNeumannCA}. Von Neumann's automaton is discussed in
    more depth later. Conway's Game of Life was thus his way of simplifying von
    Neumann's ideas, and in fact, the Game of Life is able to simulate the Game
    of Life itself and is capable of universal computation.

    The rules for Game of Life are as follows:

@@

@@

    \begin{figure}[h]
        \begin{centering}
            \begin{tabular}{cc}
                \includegraphics[width=2in]{replicator1.png} &
                \includegraphics[width=2in]{replicator2.png} \\
                %(a) first & (b) second \\[6pt]
                \includegraphics[width=2in]{replicator3.png} &
                \includegraphics[width=2in]{replicator4.png} \\
                %(c) third & (d) fourth \\[6pt]
                %\multicolumn{2}{c}{\includegraphics[width=65mm]{it} }\\
                %\multicolumn{2}{c}{(e) fifth}
            \end{tabular}
            \caption{Replicator (B1357/S1357) replicating itself over the course of
            56 generations}
        \end{centering}
    \end{figure}

@@

\section{Computational Universality}
    A system of rules, be they from a computer's instruction set, programming
    language, or cellular automaton, is called computationally universal, or
    Turing-complete, if it can be used to simulate a Turing machine.

@@

    \begin{itemize}
        \item $Q$ is the set of states
        \item $\Sigma$ is the input alphabet not containing the blank symbol,
            $\textvisiblespace$
        \item $\Gamma$ is the tape alphabet where $\textvisiblespace \in \Gamma$
            and $\Sigma \subseteq \Gamma$
        \item $\delta: Q \times \Gamma \rightarrow Q \times \Gamma \times
            \{L,R\}$ is the transition function
        \item $q_0 \in Q$ is the start state
        \item $q_{\text{accept}} \in Q$ is the accept state, and
        \item $q_{\text{reject}} \in Q$ is the reject state, with
            $q_{\text{accept}} \neq q_{\text{reject}}$
    \end{itemize}

>>>>

\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{subfig}
\usepackage{amsmath}
\usepackage{url}
\usepackage{diagbox}
\usepackage{enumitem}
\usepackage{braket}

\usepackage{tikz}
\usetikzlibrary{chains} % TM tape
\usetikzlibrary{automata,positioning}   % automaton diagram

\newenvironment{turing}[2]{
    \begin{enumerate}[leftmargin=0pt,labelsep=0pt,align=left,parsep=0pt]
        \item[$#1={}$]``\ignorespaces#2
            \begin{enumerate}[
                nosep,
                align=left,
                labelwidth=1.5em,
                label=\bfseries\arabic{*}.,
                ref=\arabic{*}
            ]}{\unskip''
            \end{enumerate}
    \end{enumerate}}

\newcommand{\bitem}{\item\hspace*{1em}\ignorespaces}

@@

    \subsection{Self-organisation}
    Second of the characteristics exhibited by this cellular automaton is
    ``self-organisation'', whereby from an initial disorderly or chaotic state
    there appear ordered structures seemingly spontaneously. This is illustrated
    in Fig. \ref{fig:self-organisation} below: one can clearly see the emergence
    of similar triangular structures in the cone snail's shell, appearing from a
    chaotic background. This starting chaos is modelled in the cellular
    automaton by randomly assigning each cell a value of 0 or 1, with each
    occurring with probability one half.

@@

    Reading the bottom row of this transition table, we get the binary number
    $01011010_{\text{bin}} = 90_{\text{dec}}$, and so we name this rule set Rule
    90. Using this convention, we may now construct our own automata based
    solely on the knowledge of its name (at least for elementary cellular
    automata). For example, take Rule 110, introduced by Wolfram 1983
    \cite{WolframRule110}. We first convert $110_{\text{dec}}$ to binary, and
    then use this to fill in the table as above:

@@

    Two technically different automata may give rise to similar patterns, and
    indeed many automata are equivalent to each other thanks to simple
    transformations of their underlying geometry. The first of these
    transformations is reflection in the vertical axis; the result of applying
    this is known as the mirror rule \cite{elementary}. For example, take Rule
    30. Each time some neighbourhood produces a live cell in Rule 30, let's take
    this neighbourhood and reflect it in the vertical plane, giving the table
    below:

@@

    The first line of the table shows the ruleset for Rule 30, while the line
    below gives that of our new ruleset, made by applying the mirror rule to
    Rule 30. Thus we get Rule 30's mirror rule, Rule $01010110_{\text{bin}}$ =
    Rule $86_{\text{dec}}$. Rules which generate the same ruleset under
    mirroring are called amphichiral, and of the 256 elementary cellular
    automata, 64 are so.

@@

    \subsection{Two dimensional cellular automata}
    Two dimensional cellular automata become a bit more interesting than their
    dimensionally-deficient brethren. Rather than having a single row of cells
    to consider at each generation, now a two-dimensional plane of cells is used
    to compute the evolution of the automaton. Whereas elementary automata
    looked only at the neighbourhood of $a_{i-1}, a_i,$ and $a_{i+1}$ when
    applying the set of rules, two-dimensional automata typically have a larger
    neighbourhood to consider. Two of the most common types of neighbourhood are
    the Moore neighbourhood and the von Neumann neighbourhood
    \cite{neighbourhoods}. The former includes the 8 cells directly surrounding
    the current cell being considered, while the latter only focuses on the four
    orthogonally ajacent cells. Occasionally, one may use the extended von
    Neumann neighbourhood, for which the next four closest orthogonally adjacent
    cells are also included.

@@

    \subsection{Game of Life and Life-like automata}
    At this point it makes sense to talk a bit more about Conway's Game of Life.
    It was invented in 1970 by the British mathematician John Conway, after
    developing an interest in them after their introduction by John von Neumann
    in the late 1940s, who aimed to discover a hypothetical machine that was
    capable of recreating itself \cite{GameOfLife}. Von Neumann was able to
    achieve this goal using a complex automaton, with 29 states and complicated
    rules to do so \cite{VonNeumannCA}. Von Neumann's automaton is discussed in
    more depth later. The Game of Life was thus Conway's way of simplifying von
    Neumann's ideas; in fact, not only can the Game of Life simulate the Game of
    Life, it is also capable of universal computation, a topic discussed more
    in-depth later. The rules for Game of Life are as follows:

@@

    On 6\textsuperscript{th} March, 2018 the first knightship was discovered by
    Adam P. Goucher, named Sir Robin \cite{Knightship}. A knightship is a
    spaceship which moves two squares horizontally for every square it moves
    vertically, as opposed to regular spaceships, which move only vertically or
    horizontally, or gliders, which move exactly diagonally. Sir Robin became
    the first new spaceship movement pattern for an elementary spaceship to be
    discovered in forty eight years \cite{fortyeight}.

@@

    \begin{figure}[h]
        \begin{centering}
            \begin{tabular}{cc}
                \includegraphics[width=2in]{replicator1.png} &
                \includegraphics[width=2in]{replicator2.png} \\
                \includegraphics[width=2in]{replicator3.png} &
                \includegraphics[width=2in]{replicator4.png} \\
            \end{tabular}
            \caption{Replicator (B1357/S1357) replicating itself over the course of
            56 generations}
        \end{centering}
    \end{figure}

@@

\section{Computational Universality}
    A system of rules, be they from a computer's instruction set, programming
    language, or cellular automaton, is called computationally universal, or
    Turing-complete, if it can be used to simulate a Turing machine. Before
    getting into the computational universality of cellular automata, we need to
    discuss the concept of Turing machines and what it means for a function to
    be computable.

@@

    \begin{itemize}
        \item $Q$ is the set of states
        \item $\Sigma$ is the input alphabet not containing the blank symbol,
            $\sqcup$
        \item $\Gamma$ is the tape alphabet, where $\sqcup \in \Gamma$
            and $\Sigma \subseteq \Gamma$
        \item $\delta: Q \times \Gamma \rightarrow Q \times \Gamma \times
            \{L,R,S\}$ is the transition function, signifying whether the tape
            head should move left, right, or stay in place, respectively
        \item $q_0 \in Q$ is the start state
        \item $q_{\text{accept}} \in Q$ is the accept state, and
        \item $q_{\text{reject}} \in Q$ is the reject state, with
            $q_{\text{accept}} \neq q_{\text{reject}}$
    \end{itemize}

    It is conventional to use $\vdash$ to signify the start of the input string,
    and an infinite string of $\sqcup$ to show the end of the input string. For
    example, the string 01011010 we would expect to be represented on the tape
    as:

    \begin{center}
        \begin{tikzpicture}
            \edef\sizetape{0.7cm}
            \tikzstyle{tmtape}=[draw,minimum size=\sizetape]
            \tikzstyle{tmhead}=[arrow box,draw,minimum size=.5cm,arrow box
            arrows={east:.25cm, west:0.25cm}]

            \begin{scope}[start chain=1 going right,node distance=-0.15mm]
                \node [on chain=1,tmtape,draw=none] {$\ldots$};
                \node [on chain=1,tmtape] {$\vdash$};
                \node [on chain=1,tmtape] {0};
                \node [on chain=1,tmtape] {1};
                \node [on chain=1,tmtape] {0};
                \node [on chain=1,tmtape] {1};
                \node [on chain=1,tmtape] {1};
                \node [on chain=1,tmtape] {0};
                \node [on chain=1,tmtape] {1};
                \node [on chain=1,tmtape] {0};
                \node [on chain=1,tmtape] {$\sqcup$};
                \node [on chain=1,tmtape] {$\sqcup$};
                \node [on chain=1,tmtape] {$\sqcup$};
                \node [on chain=1,tmtape,draw=none] {$\ldots$};
            \end{scope}
        \end{tikzpicture}
    \end{center}

    We call a set of strings Turing-recognisable \cite{Recognisable} if we can
    construct some Turing machine to recognise it, that is, every input we feed
    it that is in the language will cause the Turing machine to eventually move
    to the accept state. Notice, however, that while there are two
    ``destination'' states, $q_{\text{accept}}$ and $q_{\text{reject}}$, there
    is a third possible outcome: that our machine does not halt, but instead
    loops somewhere in its proceedings infinitely. Thus we call a set of strings
    decidable if our Turing machine always halts for its inputs, that is, for
    any input we feed it we are guaranteed to land in either $q_{\text{accept}}$
    or $q_{\text{reject}}$.

    As an example of a Turing machine in action, suppose we want to construct a
    one to accept the language $L = \{w\#w | w \in \{0,1\}\ast\}$.  This is not
    Context Free \cite{nonCFL}, hinting at the power behind Turing Machines with
    regard to computability. We construct the machine informally as follows:

    \begin{turing}{M}{on input $w$:}
        \item Find a decomposition of $w$ into $w_1 \# w_2$
        \item Start at the first symbol of $w_1$, replace the symbol with a
            cross, and skip forward to the first symbol of $w_2$.
            \bitem if this symbol is the same as the one you just crossed off,
            cross it off and rewind to the symbol on the right of the first
            symbol of $w_1$
            \bitem else reject
        \item Repeat this process, each time skipping forward and rewinding so
            that you are inspecting the first symbol to the right of a cross
            each time
        \item If you end up inspecting a cell with the symbol $\#$, and the
            symbols either side of it are crosses, accept
    \end{turing}

    The final thing to cover in order to understand how cellular automata such
    as the Game of Life can be used to simulate a Turing Machine is the concept
    of a Universal Turing Machine. A Universal Turing Machine is one which
    recognises the language

    $$A_{TM} = \{\braket{M,w} | M \text{ is a Turing Machine and } M
    \text{ accepts } w\} $$

    The $\braket{M,w}$ notation is used to denote the encoding of the Turing
    Machine $M$ and its input $w$ in a way that some Turing Machine would be
    able to understand and use as input. In essence, Universal Turing Machines
    are those which are capable of simulating any other Turing Machine just from
    its description and an input, and it is from this angle that we will look at
    the universality of certain cellular automata.

    \begin{figure}
        \centering
        \begin{tikzpicture}[
                ->,
                node distance=3cm,
                on grid,
                auto]
            \node[state,initial]    (q1) []                     {$q_1$};
            \node[state]            (q2) [above right of=q1]    {$q_2$};
            \node[state]            (q3) [right of=q2]          {$q_3$};
            \node[state]            (q4) [below right of=q1]    {$q_4$};
            \node[state]            (q5) [right of=q4]          {$q_5$};
            \node[state]            (q6) [above right of=q5]    {$q_6$};
            \node[state]            (q7) [right of=q6]          {$q_7$};
            \node[state,accepting]  (qa) [right of=q7]          {$q_a$};
            \node[state,accepting]  (qr1) [right of=q3]         {$q_r$};
            \node[state,accepting]  (qr2) [right of=q5]         {$q_r$};

            \path[->]
            (q1)
            edge [loop above] node [align=center] {\tiny$\vdash \rightarrow R$
            \\ \tiny$\times \rightarrow R$} ()
            edge  node  [swap] {\tiny$0 \rightarrow \times, R$} (q2)
            edge  node  {\tiny$1 \rightarrow \times, R$} (q4)
            (q2)
            edge [loop above] node [align=center] {\tiny$0 \rightarrow R$ \\
            \tiny$1 \rightarrow R$} ()
            edge node {\tiny$\# \rightarrow R$} (q3)
            (q3)
            edge [loop above] node {\tiny$\times \rightarrow R$} ()
            edge node [swap] {\tiny$0 \rightarrow \times, L$} (q6)
            edge node {\tiny$1 \rightarrow \times, S$} (qr1)
            (q4)
            edge [loop below] node [align=center] {\tiny$0 \rightarrow R$ \\
            \tiny$1 \rightarrow R$} ()
            edge node {\tiny$\# \rightarrow R$} (q5)
            (q5)
            edge [loop below] node {\tiny$\times \rightarrow R$} ()
            edge node {\tiny$0 \rightarrow \times, S$} (qr2)
            edge node {\tiny$1 \rightarrow \times, L$} (q6)
            (q6)
            edge [loop above] node [align=center] {\tiny$0 \rightarrow L$ \\
            \tiny$1 \rightarrow L$} ()
            edge [loop below] node [align=center] {\tiny$\times \rightarrow L$} ()
            edge [bend left] node {\tiny$\# \rightarrow L$} (q7)
            edge node [swap] {\tiny$\vdash \rightarrow R$} (q1)
            (q7)
            edge [bend left] node [align=center] {\tiny$0 \rightarrow L$ \\
            \tiny$1 \rightarrow L$} (q6)
            edge node {\tiny$\times \rightarrow S$} (qa);
        \end{tikzpicture}
        \caption{A possible Turing Machine to recognise $L$}
        \label{TM:L}
    \end{figure}

    \subsection{Computability and Universality of Cellular Automata}
    A Turing Machine may compute a function by starting with the input to the
    function on the tape and halting with the output of the function on the
    tape. A function $f : \Sigma \ast \rightarrow \Sigma \ast$ is computable if
    some Turing Machine $M$ halts, for every input $w$, with just the result
    $f(w)$ on the tape \cite{Computable}. Thus, any model we can devise which
    could simulate the run of a Turing Machine on some input would be capable of
    computing all the functions that the Turing Machine could, and hence such a
    model would be Turing complete.
